---
title: "Data wrangling I: dplyr verbs"
output:
  rmarkdown::html_document:
    theme: lumen
    toc: true
    toc_float: true
---


NOTE: You will not be able to knit this document until you fill in some blanks below.


\
\



# Why R / RStudio? (A reminder)

Beyond some simple calculations, you can't do data analysis without software. So why R / RStudio?

- It's free.
- It has a huge online community (thus finding help is easier than with other software).
- Fun fact: RStudio was started by Macalester alum JJ Allaire and beta-tested at Mac!
- It's used outside academia, and not just by statisticians.    
    - [BBC uses R.](https://bbc.github.io/rcookbook/)    
    - Minnesota Public Radio journalist [David Montgomery](http://dhmontgomery.com/portfolio/) uses R for data analysis and [visualizations](https://github.com/dhmontgomery/personal-work/tree/master/theme-mpr).
    - [Ahmadou Dicko](https://www.rstudio.com/resources/rstudioglobal-2021/humanitarian-data-science-with-r/) discusses how humanitarians are using R to create "life saving data products."
    - [Shelmith Kariuki](https://www.rstudio.com/resources/rstudioglobal-2021/rkenyacensus-package/) discusses how the Kenyan government shared its census data using R, to support policy making and development.
    


\
\


# dplyr verbs

We can know all the theory and fancy algorithms in the world. But if we can't **wrangle our data**, we won't get anywhere in a data analysis. Today we'll explore how to wrangle **tidy data**^[Each column is a variable and each row is a case. See [Wickham, Tidy Data](https://www.jstatsoft.org/article/view/v059i10) and 
[Wickham and Grolemund, Tidy Data](http://r4ds.had.co.nz/tidy-data.html) for more.] using the `dplyr` package which, along with `ggplot2`, is part of the broader **tidyverse**. In the `dplyr` grammar, there are 6 main data transformation _verbs_ or actions:

\
\



verb            action                                            example 
--------------- ------------------------------------------------- ---------------------------
`select()`      **select** a subset of _columns_                  `select(x, y)`, `select(-x)`
`mutate()`      create or **mutate** a new variable / _column_    `mutate(x = ___, y = ___)`
`summarize()`   calculate a numerical *summary* of a _column_     `summarize(mean(x), median(y))`
`group_by()`    **group** the _rows_ by a specified _column_      `group_by(x) %>% summarize(mean(y))`
`arrange()`     **arrange** the _rows_                            `arrange(x)`, `arrange(desc(x))`   
`filter()`      **filter** out / obtain a subset of _rows_        `filter(x == __, y > __)`  



\
\


We can apply one or more verbs through a series of __pipes__ (`%>%`):    

```{r eval = FALSE}
# Apply 1 verb to my_data
my_data %>% 
  verb_1(___)

# Apply 2 verbs to my_data
my_data %>% 
  verb_1(___) %>% 
  verb_2(___)
```



\
\



# Exercises


The fivethirtyeight article
[The Ultimate Halloween Candy Power Ranking](http://fivethirtyeight.com/features/the-ultimate-halloween-candy-power-ranking/) analyzed data from [this experiment](http://walthickey.com/2017/10/18/whats-the-best-halloween-candy/) which presented subjects with a series of head-to-head candy match-ups and asked them to indicate which candy they preferred. You can load these data from the `fivethirtyeight` package:

```{r message = FALSE, warning = FALSE}
# Load data
library(fivethirtyeight)
data("candy_rankings")

# Store under a shorter name
candy <- candy_rankings

head(candy)
nrow(candy)
```

And load the `dplyr` package which contains the data wrangling functions:

```{r warning = FALSE, message = FALSE}
# Load the dplyr package
library(dplyr)
```


\


We'll wrangle these data below.  In each exercise, first identify which `dplyr` verbs will be useful.


\
\



## Exercise 1: Sort the candies


(`select` columns, `mutate` a column, `summarize` a column, `arrange` rows, `filter` rows, `group_by` groups)    

```{r}
# Sort the candy from least to most popular
#Sort=arrange
candy %>% 
  arrange(winpercent) %>%
  select(competitorname,winpercent)
```

```{r}
# Show the 6 least popular candies
candy %>% 
  arrange(winpercent) %>% 
  head()
```

```{r}
# Show the 6 most popular candies
candy %>% 
  arrange(desc(winpercent)) %>% 
  head()
```


\
\



## Exercise 2: Find the candies

(`select` columns, `mutate` a column, `summarize` a column, `arrange` rows, `filter` rows, `group_by` groups)    

\


NOTE: R recognizes the following qualifiers:

`<, <=, ==, >, >=, != (not equal), %in%`



```{r}
# Show the data for Snickers
candy %>% 
  filter(competitorname == "Snickers")
```

```{r}
# Create a data set with only the chocolate candies
# Show the first 6 entries
candy %>% 
  filter(chocolate == TRUE) %>% 
  head()
```

```{r}
# Create a data set of the candies that won less than 45% of their match-ups
# Show the first 6 entries
candy %>% 
  filter(winpercent < 45) %>% 
  head()
```

```{r}
# In 3 lines, show the chocolate candies that won less than 45% of their match-ups

candy %>%
  filter(chocolate==TRUE) %>% 
  filter(winpercent<45)
```

```{r}
# In 2 lines, show the chocolate candies that won less than 45% of their match-ups
candy %>% 
  filter(chocolate==TRUE,winpercent<45)
```
    


\
\



## Exercise 3: Define new variables

(`select` columns, `mutate` a column, `summarize` a column, `arrange` rows, `filter` rows, `group_by` groups)    

Create and save two new variables in the `candy` dataset:    

- Redefine `sugarpercent` from 0-1 scale to the 0-100 scale

- Define a `choc_peanut` variable which identifies candies that contain both chocolate & peanuts/almonds

We'll do this in two steps (why?!?):

- Convince ourselves that our code is doing what we want it to.
- Then save the new variables in the `candy` dataset.


```{r}
# Play around first
#Mutate is creating a new column, or like a new "variable"
candy %>% 
  mutate(sugarpercent = sugarpercent*100,
      choc_peanut = (chocolate == TRUE & peanutyalmondy == TRUE)) %>% 
  head() %>% 
  select(competitorname,chocolate,peanutyalmondy,choc_peanut,sugarpercent)
```

```{r}
# Then store the results
# (How does this code differ from that above?) We don't need the head function since we're storing all of it. Make sure your code is right before overwriting your data.
#Also, if you run the chunk multiple times, it overwrites it multiple times! This means sugarpercent keeps getting multiplied by 100!
candy <- candy %>% 
  mutate(sugarpercent = sugarpercent,
      choc_peanut = (chocolate == TRUE & peanutyalmondy == TRUE))
```
    



\
\




## Exercise 4: Calculate some stats

(`select` columns, `mutate` a column, `summarize` a column, `arrange` rows, `filter` rows, `group_by` groups)    

```{r}
# Calculate the median and maximum winpercent
candy %>% 
  summarize(median(winpercent), max(winpercent))
```

```{r}
# Calculate the median and maximum winpercent for chocolate vs non-chocolate candies
# Hint: we are calculating summaries for different *groups*
#Wanting to specifically look at chocolate
candy %>% 
  group_by(chocolate) %>% 
  summarize(median(winpercent),max(winpercent))
#Order of operations matters, need to group first
```

```{r}
# Group the candies by chocolate vs non-chocolate candies
# What do you get??
candy %>% 
  group_by(chocolate)
```



\
\



## Exercise 5: Pipe series

(`select` columns, `mutate` a column, `summarize` a column, `arrange` rows, `filter` rows, `group_by` groups)    

When we have a *series* of pipes, it's important to ask: Does the order of operations impacts the result? If not, is one order "better" than another?

```{r}
# Arrange the fruity candies from most to least popular
candy %>% 
  filter(fruity==TRUE) %>% 
  arrange(desc(winpercent)) %>% 
  head() %>% 
  select(competitorname,fruity,winpercent)

# Among chocolate candies, what's the median winpercent?
candy %>% 
  filter(chocolate==TRUE) %>% 
  summarize(median(winpercent))

# Among chocolate candies, how does the median winpercent compare if the candy is also peanutyalmondy?
candy %>% 
  filter(chocolate==TRUE) %>% 
  group_by(peanutyalmondy) %>% 
  summarize(median(winpercent))

```
    
    
