---
title: "Homework 3: Confidence and prediction intervals"
output:
  rmarkdown::html_document:
    theme: lumen
    toc: true
    toc_float: true
date: "2023-06-22"
---
9.2 Toward hypothesis testing
Using the high_peaks data on hiking trails in the Adirondack mountains of northern New York state, we’ll explore whether the time it takes to complete each hike is associated with various features: a hike’s highest elevation, length, ascent, and difficulty rating.
```{r}
# Load packages
library(ggplot2)
library(dplyr)

# Import data
peaks <- read.csv("https://raw.githubusercontent.com/ajohns24/data/main/high_peaks.csv")
```

2. Constructing a confidence interval
Consider the relationship between y, the time it takes to complete a hike (in hours), by x, its highest elevation (in feet):

y=β0+β1x+ε

    Estimate this model using our sample data.
    Use the 68-95-99.7 Rule along with information from the model table to calculate an approximate 95% CI for β1, the actual elevation coefficient.
    Then check your work using confint.
    
3. CI Interpretation
a) What can we conclude from the 95% CI for β1?

    Our sample data establishes a discernible negative association between hiking time and elevation.
    Our sample data establishes a discernible positive association between hiking time and elevation.
    Our sample data does not establish a discernible association between hiking time and elevation.

b) Alternatively (yet equivalently), answer this question using the confidence bands around our sample model trend. NOTE: These confidence bands are reminiscent of our sampling distribution simulations which produced a range of model trends.
```{r}
ggplot(peaks, aes(x = elevation, y = time)) + 
  geom_point() + 
  geom_smooth(method = "lm")
```

4. Another predictor
Next, let’s use our data to make inferences about the relationship between the time it takes to complete a hike in hours (y), by the hike’s vertical ascent, or the total change in elevation in feet (x).

    Use your sample to estimate this model.
    Calculate a 95% CI for the population ascent coefficient.
    How can we interpret this CI?
        We’re 95% confident that for every 1 foot increase in ascent, the completion time for all hikes in the population typically increases by somewhere between 0.0009 and 0.003 hours. (Or, increases somewhere between 0.9 and 3 hours for every 1000 foot increase in ascent.)
        We’re 95% confident that for every 1 foot increase in ascent, the completion time for all hikes in our sample typically increases by somewhere between 0.0009 and 0.003 hours. (Or, increases somewhere between 0.9 and 3 hours for every 1000 foot increase in ascent.)
        For 95% of all hikes, completion time will increase peaks size will increase by somewhere between 0.0009 and 0.003 hours for every extra 1 foot increase in ascent.

5. Interpretation
a) What can we conclude from the 95% CI for β1?

    Our sample data establishes a discernible negative association between hiking time and ascent.
    Our sample data establishes a discernible positive association between hiking time and ascent.
    Our sample data does not establish a discernible association between hiking time and ascent.

b) Alternatively (yet equivalently), plot the confidence bands around our sample model trend and use these to answer the question.

6. Statistically discernible vs practically meaningful
In the previous example, we concluded that there’s a statistically discernible association between hiking time and ascent. This indicates that an association exists. It does not necessarily mean that the association is meaningful, i.e. practically significant.

It’s important to follow up a technical confidence interval with some context. To this end, do you think that the association between hiking time and ascent is also practically significant? Mainly, in the context of hiking, is the magnitude of the association (i.e. an increase between 0.9 and 3 hours for every 1000 foot increase in ascent) actually meaningful?

7. Two predictors
Next let’s explore the relationship between the time it takes to complete a hike, by the hike’s ascent and its length (in miles).

    Estimate the model of this relationship.
    Construct a 95% CI for the actual ascent coefficient.
    What can we conclude from the 95% CI?
        There’s a discernible association between hiking time and ascent.
        There’s not a discernible association between hiking time and ascent.
        When controlling for hike length, there’s a discernible association between hiking time and ascent.
        When controlling for hike length, there’s not a discernible association between hiking time and ascent.

8. Reflection
When considered alone, vertical ascent was a discernible predictor of hiking time. Yet when considered along with hike length, it wasn’t.

    Explain why this happened.
    Support your explanation with some new graphical or numerical evidence.

9. Conclusions
Reconsider the model with both ascent and length predictors. If we already have vertical ascent in our model, is hike length a useful predictor of hiking time? Why or why not?

9.3 Prediction Intervals

10. Model prediction
Consider a model of hiking time (in hours) vs length (in miles):
```{r}
lm_spec <- 
  linear_reg() %>% 
  set_mode("regression") %>% 
  set_engine("lm") 

model_4 <- lm_spec %>% 
  fit(time ~ length, data = peaks)

model_4 %>% 
  tidy()
```
There are 2 types of predictions we can make for, say, a hiking length of 10 miles:

    Predict the average hiking time for all 10-mile hikes in the world.
    Predict the hiking time for the “Minnesota trail”, a specific 10-mile hike.

The values of the two predictions are the same:
  time = 2.05 + 0.684*10 = 8.89 hours

However, the potential error in these predictions differs. Check in with your intuition: Is there more error in trying to predict the average hiking time of all 10-mile hikes in the world, or the exact hiking time of the Minnesota trail, a specific 10-mile hike? Explain your reasoning.

11. Confidence intervals and confidence bands
a) Check your intuition. Calculate and report the 95% confidence interval for the average hiking time of all 10-mile hikes:
```{r}
# Just the estimated average
model_4 %>% 
  predict(new_data = data.frame(length = 10))

# A confidence interval for the estimated average
model_4 %>% 
  predict(
    new_data = data.frame(length = 10), 
    type = "conf_int", 
    level = 0.95)
```

b) What's the best interpretation of this interval

    Among the 10-mile long hikes in our sample, we’re 95% confident that the average hiking time is in this interval.
    Among all 10-mile long hikes in the world, we’re 95% confident that the average hiking time is in this interval.
    We’re 95% confident that the hiking time for the Minnesota trail is in this interval.

c) We can visualize the confidence interval for the average hiking length of all hikes at any common length (not just 10 miles) by drawing confidence bands around the model. The vertical line drawn at a length of 10 miles reflects the interval you calculated above. NOTE: What have we changed in the syntax?
```{r}
ggplot(peaks, aes(x = length, y = time)) + 
  geom_point() + 
  geom_smooth(method = "lm") + 
  geom_segment(aes(x = 10, xend = 10, y = 8.35, yend = 9.43), color = "red")
```

12. Prediction intervals
a) Calculate and report the 95% prediction interval (PI) for the hiking time for the Minnesota trail, a specific 10-mile hike. NOTE: What changed in the syntax?
```{r}
# Just the prediction
model_4 %>% 
  predict(new_data = data.frame(length = 10))

# A prediction interval
model_4 %>% 
  predict(
    new_data = data.frame(length = 10), 
    type = "pred_int", 
    level = 0.95)
```

b) How can we interpret this interval?

    Among the 10-mile long hikes in our sample, we’re 95% confident that the average hiking time is in this interval.
    Among all 10-mile long hikes in the world, we’re 95% confident that the average hiking time is in this interval.
    We’re 95% confident that the hiking time for the Minnesota trail is in this interval.

13. Prediction vs confidence bands
We can visualize the prediction interval for the hiking time of a trail at any length by drawing prediction bands. This requires some messy syntax:
```{r}
# Calculate and store prediction intervals for every length value
predictions_1 <- model_4 %>% 
  predict(new_data = peaks, type = "pred_int") %>% 
  bind_cols(peaks)

# Check it out
predictions_1 %>% 
  dplyr::select(peak, length, time, .pred_lower, .pred_upper) %>% 
  head()
```
```{r}
# Plot regression line with prediction bands
ggplot(predictions_1, aes(x = length, y = time)) + 
  geom_point() + 
  geom_smooth(method = 'lm', color = "blue") + 
  geom_ribbon(aes(y = (.pred_lower + .pred_upper) / 2, ymin = .pred_lower, ymax = .pred_upper, fill = 'prediction'), alpha = 0.2)
```
a) Do the confidence bands (darker gray) capture our uncertainty about the hiking time for the AVERAGE HIKE or for INDIVIDUAL / SPECIFIC HIKES?

b) Do the prediction bands (lighter gray) capture our uncertainty about the hiking time for the AVERAGE HIKE or for INDIVIDUAL / SPECIFIC HIKES?

c) Which are wider, intervals for the average hiking time for all hikes of a common length OR intervals for the exact hiking time for a specific, individual hike at that length? Explain why this makes intuitive sense.

14. Narrowing
Though it’s not as noticeable with the prediction bands, these and the confidence bands are always the most narrow at the same point – in this case at a length of 12.57391 miles. What other meaning does this value have? Provide some proof and explain why it makes intuitive sense that the bands are narrowest at this point.

15. Extra practice
Reconsider the model of hiking time vs ascent, model_2.
a) Construct and interpret a 95% prediction interval for the hiking time for the Minnesota Trail, which has a vertical ascent of 1800 feet.

b) Construct and interpret a 95% confidence interval for the average hiking time among all hikes that have an 1800-foot vertical ascent.

9.4 Data wrangling

16. Billboard
The billboard data in the tidyr package includes “song rankings for Billboard top 100 in the year 2000”.
```{r}
library(tidyr)
data(billboard)
```
For only the first 12 songs in this dataset, plot the song’s ranking over time, from weeks 1 to 76. Give each song its own facet, labeled by track (not artist).


9.5 Solutions

3. CI Interpretation
a) Our sample data does not establish a discernible association between hiking time and elevation.
b) The bands include 0-slope lines

4. Another predictor
We’re 95% confident that for every 1 foot increase in ascent, the completion time for all hikes in the population typically increases by somewhere between 0.0009 and 0.003 hours. (Or, increases somewhere between 0.9 and 3 hours for every 1000 foot increase in ascent.)

5. Interpretation
a) Our sample data establishes a discernible positive association between hiking time and ascent.
b) The bands do not include any 0-slope lines.

6. Statistically discernible vs practically meaningful
I think so.

7. Two predictors
When controlling for hike length, there's not a discernible association between hiking time and ascent.

8. Reflection
Multicollinearity. When length isn’t in the model, vertical ascent is a discernible predictor of hiking time. Yet the plot below confirms that hikes with greater ascent also tend to be longer (ascent and length are correlated). Thus if we already know a hike’s length, then its vertical ascent doesn’t provide a discernible amount of additional information about hiking time.

9. Conclusions
Yes. Its CI is entirely and meaningfully above 0.

10. Model prediction
Your intuition will vary. Your intuition might be right or wrong, the important thing is to pause and reflect.

11. Confidence intervals and confidence bands
b) Among all 10-mile long hikes in the world, we’re 95% confident that the average hiking time is in this interval.

12. Prediction intervals
a) NOTE: we changed conf_int to pred_int
b) We’re 95% confident that the hiking time for the Minnesota trail is in this interval.

13. Prediction vs confidence bands
a) average hike
b) individual / specific hikes
c) Intervals for the exact hiking time for a specific, individual hike are wider. It’s tougher to anticipate the exact behavior of an individual data point (it could be an outlier!), and easier to anticipate the average behavior of a group of data points (where any outliers are averaged out).

14. Narrowing
This is the average hiking length. Why’s there less error here? Conceptually, it’s easier to anticipate the hiking time (y) for hikes of the typical length. Further, there are more data points, hence information, in the middle here than at the extremes of hiking length.