---
title: "Mini-project"
output:
  rmarkdown::html_document:
    theme: lumen
    toc: true
    toc_float: true
date: "2023-08-27"
---
15 Mini-project

15.1 Data

15.1.1 Option 1: Weather forecasting
Use the weatherAUS data in the rattle package to play meteorologist. For example, can you predict tomorrow’s temperature? What about whether or not it rains tomorrow?
```{r}
library(rattle)
data(weatherAUS)
```
NOTE: In your console, not Rmd, you can access a data dictionary by typing ?weatherAUS.

15.1.2 Option 2: Hotel cancellations
Use the data posted by TidyTuesday to explore hotel cancellations.
```{r}
hotels <- read.csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-02-11/hotels.csv")
```

15.1.3 Option 3: Basketball
Use data from NBA shot logs to predict whether or not a player will make their shot.
```{r}
bball <- read.csv("https://raw.githubusercontent.com/ajohns24/data/main/shot_logs.csv")
```

15.2 Step 1
Understand the basic data that is available to you.

    Where does the data come from? How was it collected?
    Is there a data dictionary?
    How much data is there?
    What variables do you have and how are these structured?

15.3 Step 2
Visualize and describe the variables that seem most interesting or relevant. Choose a final set of visualizations that help tell a complete story.

15.4 Step 3
Now that you have a sense of what you’re working with, formulate a more specific research question.

15.5 Step 4
Analyze the data related to the research question, starting from simple analyses to more complex ones. For example, look at more visualizations, build some models, evaluate some models, and decide which models to use that help tell a complete story.

15.6 Step 5
Interpret your findings, refine your research question, and return to step 4. Iterate!